Evaluating FAIR Digital Object as a distributed object system

### 2023-01-25 latex/main.tex
latex syntax fun

In other discourse, Linked Data \cite{19s3yyxDn} has been seen as an established set of principles based on Semantic Web technologies that can achieve the vision of the FAIR principles (\cite{EJDjHrUv}; \cite{GJHYN6xW}). Yet regular researchers and developers of emerging platforms for computation and data management are reluctant to adapt such a FAIR Linked Data approach fully \cite{jHwttlTs}, opting instead for custom in-house models and JSON-derived formats from RESTful Web services (\cite{99MS5xE0}; \cite{J2H8yssV}). While such focus on simplicity gives rapid development and highly specialised services, it raises wider concerns on interoperability (\cite{DNTCUjpC}; \cite{Pa67pUtR}).

One challenge that may, perhaps counter-intuitively, steer developers towards a not-invented-here mentality (\cite{uxWjmlio}; \cite{v0WwFHgT}) when exposing their data on the Web is the heterogeneity and apparent complexity of Semantic Web approaches themselves \cite{5V3GqSzR}.

@@

The creation of RDF-based knowledge graphs grew particularly in fields like bioinformatics, e.g.~for describing genomes and proteins (\cite{bMSxa8XW}; \cite{Qlrlr2TZ}). In theory, the use of RDF by the life sciences would enable interoperability between the many data repositories and support combined views of the many aspects of bio-entities -- however in practice most institutions ended up making their own ontologies and identifiers, for what to the untrained eye would mean roughly the same. One can argue that the toll of adding the semantic logic system of rich ontologies meant that small, but fundamental, differences in opinion (e.g.~should a \emph{gene identifier} signify which protein a DNA sequence would make, or just the particular DNA sequence letters, or those letters as they appear in a particular position on a human chromosome?) lead to large differences in representational granularity, and thus needed different identifiers.

@@

With these trends, an emerging problem was that adopters of the Semantic Web primarily utillised it as a set of graph technologies, with little consideration to existing Web resources. This meant that links stayed mainly within a single information system, with little URI reuse even with large term overlaps \cite{9CdJnsoJ}. Just like \emph{link rot} affect regular Web pages and their citations from scholarly communication \cite{AGHR9oyr}, for a majority of described RDF resources in the \href{https://lod-cloud.net/}{Linked Open Data} (LOD) Cloud's gathering of more than thousand datasets, unfortunately they do not actually link to (still) downloadable (\emph{dereferenceable}) Linked Data \cite{5SuwLNdE}. Another challenge facing potential adopters is the plethora of choices, not just to navigate, understand and select to reuse the many possible vocabularies and ontologies \cite{TUmA4KQT} , but also technological choices on RDF serialisation (at least \href{https://www.w3.org/TR/rdf11-primer/\#section-graph-syntax}{7 formats}), type system (RDFS \cite{ZwAcGQKY}, OWL \cite{1p4IWJpI}, OBO \cite{1DalZuKct}, SKOS \cite{15gQDya5B}), hash vs slash, HTTP status codes and PID redirection strategies \cite{rnzJ4khD}.

@@

Recently there has been a renewed emphasis to improve the \emph{Developer Experience} (\cite{BFMoe8FC}, 2018}) for consumption of Linked Data, for instance RDF Shapes (expressed in SHACL \cite{GDrm9dgX} or ShEx \cite{YUzogaV}) can be used to validate RDF Data (\cite{ZeOAPNHa}; \cite{S5BI6Mwx}) before consuming it programmatically, or reshaping data to fit other models. While a varied set of tools for Linked Data consumptions have been identified, most of them still require developers to gain significant knowledge of the underlying technologies, which hampers adaption by non-LD experts \cite{o596f2JM}, which then tend to prefer non-semantic two-dimensional formats such as CSV files.

A valid concern is that the Semantic Web research community has still not fully embraced the Web, and that the ``final 20\%'' engineering effort is frequently overlooked in favour of chasing new trends such as Big Data and AI, rather than making powerful Linked Data technologies available to the wider groups of Web developers \cite{jHwttlTs}. One bridging gap here by the Linked Data movement has been ``linked data by stealth'' approaches such as structured data entry spreadsheets powered by ontologies \cite{MhsCt4P3}, the use of Linked Data as part of REST Web APIs \cite{eS6a6tlz} , and as shown by the big uptake by publishers to annotate the Web using schema.org Bernstein, Hendler \& Noy \cite{117tX9MEG}, with vocabulary use patterns documented by copy-pastable JSON-LD examples, rather than by formalised ontologies or developer requirements to understand the full Semantic Web stack.

@@

\begin{itemize}
\tightlist
\item
  Linked Data in general is strong on metadata indicators, but LDP approach is weak as it has little metadata guidance
\item
  FDO/DOIP are stronger on identifier indicators
\item
  Indicators on standard protocols (RDA-A1-04M, RDA-A1-04D, RDA-A1.1-01M, RDA-A1.1-01D) favour LDP's mature standards (HTTP, URI) -- the DOIPv2 specification \cite{13TcbsZF6} has currently only a couple of implementations and is expressed informally. The underlying Handle system for PIDs is arguably mature and commonly used by researchers (this article alone references about 80 DOIs), however DOIs are more commonly accessed as HTTP redirects through resolvers like \url{https://doi.org/} and \url{http://hdl.handle.net/} rather than the Handle protocol.
\item
  RDA-A1-02M and RDA-A1-02D highlights access by manual intervention, which is common for http/https URIs, but also using above PID resolvers for DOIP implementation \href{https://www.cordra.org/}{CORDRA} (e.g.~\url{https://hdl.handle.net/21.14100/90ec1c7b-6f5e-4e12-9137-0cedd16d1bce}), yet neither LDP, FDO nor DOIP specifications recommends human-readable representations to be provided
\item
  Neither DOIP nor LDP require license to be expressed (RDA-R1.1-01M, RDA-R1.1-02M, RDA-R1.1-03M), yet this is crucial for re-use and machine actionability of FAIR data and metadata to be legal
\item
  Machine-understandable types, provenance and data/metadata standards (RDA-R1.1-03M RDA-R1.3-02M, RDA-R1.3-02M, RDA-R1.3-02D) are important for machine actionability, but are currently unspecified for FDOs. \cite{126uxr5pI} explores possible machine-readable FDO types, however the type systems themselves have not yet been formalised. Linked Data on the other side have too many semantic and syntactic type systems, making it difficult to write consistent clients.
\item
  Indicators for FAIR data are weak for either approach, as too much reliance is put on metadata. For instance in Linked Data, given a URL of a CSV file, what is its persistant identifier or license information? FAIR Signposting \cite{snykkm7R} can improve findability of metadata using HTTP Link relations, which enable an FDO-like overlay for any HTTP resource. In DOIP, responses for bytestreams can include the data identifier: if that is a PID (not enforced by DOIP), its metadata is accessible.
\item
  Resolving FDOs via Handle PIDs to the corresponding DOIP server is currently undefined by FDO and DOIP specifications. \texttt{0.TYPE/DOIPServiceInfo} lookup is only possible once DOIP server is known. g
\end{itemize}

>>>>

In other discourse, Linked Data \cite{19s3yyxDn} has been seen as an established set of principles based on Semantic Web technologies that can achieve the vision of the FAIR principles \cite{EJDjHrUv,GJHYN6xW}. Yet regular researchers and developers of emerging platforms for computation and data management are reluctant to adapt such a FAIR Linked Data approach fully \cite{jHwttlTs}, opting instead for custom in-house models and JSON-derived formats from RESTful Web services \cite{99MS5xE0,J2H8yssV}). While such focus on simplicity gives rapid development and highly specialised services, it raises wider concerns on interoperability \cite{DNTCUjpC,Pa67pUtR}.

One challenge that may, perhaps counter-intuitively, steer developers towards a not-invented-here mentality \cite{uxWjmlio,v0WwFHgT} when exposing their data on the Web is the heterogeneity and apparent complexity of Semantic Web approaches themselves \cite{5V3GqSzR}.

@@

The creation of RDF-based knowledge graphs grew particularly in fields like bioinformatics, e.g.~for describing genomes and proteins \cite{bMSxa8XW,Qlrlr2TZ}. In theory, the use of RDF by the life sciences would enable interoperability between the many data repositories and support combined views of the many aspects of bio-entities -- however in practice most institutions ended up making their own ontologies and identifiers, for what to the untrained eye would mean roughly the same. One can argue that the toll of adding the semantic logic system of rich ontologies meant that small, but fundamental, differences in opinion (e.g.~should a \emph{gene identifier} signify which protein a DNA sequence would make, or just the particular DNA sequence letters, or those letters as they appear in a particular position on a human chromosome?) lead to large differences in representational granularity, and thus needed different identifiers.

@@

With these trends, an emerging problem was that adopters of the Semantic Web primarily utillised it as a set of graph technologies, with little consideration to existing Web resources. This meant that links stayed mainly within a single information system, with little URI reuse even with large term overlaps \cite{9CdJnsoJ}. Just like \emph{link rot} affect regular Web pages and their citations from scholarly communication \cite{AGHR9oyr}, for a majority of described RDF resources in the \href{https://lod-cloud.net/}{Linked Open Data} (LOD) Cloud's gathering of more than thousand datasets, unfortunately they do not actually link to (still) downloadable (\emph{dereferenceable}) Linked Data \cite{5SuwLNdE}. Another challenge facing potential adopters is the plethora of choices, not just to navigate, understand and select to reuse the many possible vocabularies and ontologies \cite{TUmA4KQT}, but also technological choices on RDF serialisation (at least \href{https://www.w3.org/TR/rdf11-primer/\#section-graph-syntax}{7 formats}), type system (RDFS \cite{ZwAcGQKY}, OWL \cite{1p4IWJpI}, OBO \cite{1DalZuKct}, SKOS \cite{15gQDya5B}), hash vs slash, HTTP status codes and PID redirection strategies \cite{rnzJ4khD}.

@@

Recently there has been a renewed emphasis to improve the \emph{Developer Experience} \cite{BFMoe8FC} for consumption of Linked Data, for instance RDF Shapes (expressed in SHACL \cite{GDrm9dgX} or ShEx \cite{YUzogaV}) can be used to validate RDF Data \cite{ZeOAPNHa,S5BI6Mwx} before consuming it programmatically, or reshaping data to fit other models. While a varied set of tools for Linked Data consumptions have been identified, most of them still require developers to gain significant knowledge of the underlying technologies, which hampers adaption by non-LD experts \cite{o596f2JM}, which then tend to prefer non-semantic two-dimensional formats such as CSV files.

A valid concern is that the Semantic Web research community has still not fully embraced the Web, and that the ``final 20\%'' engineering effort is frequently overlooked in favour of chasing new trends such as Big Data and AI, rather than making powerful Linked Data technologies available to the wider groups of Web developers \cite{jHwttlTs}. One bridging gap here by the Linked Data movement has been ``linked data by stealth'' approaches such as structured data entry spreadsheets powered by ontologies \cite{MhsCt4P3}, the use of Linked Data as part of REST Web APIs \cite{eS6a6tlz}, and as shown by the big uptake by publishers to annotate the Web using schema.org \cite{117tX9MEG}, with vocabulary use patterns documented by copy-pastable JSON-LD examples, rather than by formalised ontologies or developer requirements to understand the full Semantic Web stack.

@@

\begin{itemize}
\tightlist
\item
  Linked Data in general is strong on metadata indicators, but LDP approach is weak as it has little metadata guidance
\item
  FDO/DOIP are stronger on identifier indicators
\item
  Indicators on standard protocols (RDA-A1-04M, RDA-A1-04D, RDA-A1.1-01M, RDA-A1.1-01D) favour LDP's mature standards (HTTP, URI) -- the DOIPv2 specification \cite{13TcbsZF6} has currently only a couple of implementations and is expressed informally. The underlying Handle system for PIDs is arguably mature and commonly used by researchers (this article alone references about 80 DOIs), however DOIs are more commonly accessed as HTTP redirects through resolvers like \url{https://doi.org/} and \url{http://hdl.handle.net/} rather than the Handle protocol.
\item
  RDA-A1-02M and RDA-A1-02D highlights access by manual intervention, which is common for http/https URIs, but also using above PID resolvers for DOIP implementation \href{https://www.cordra.org/}{CORDRA} (e.g.~\url{https://hdl.handle.net/21.14100/90ec1c7b-6f5e-4e12-9137-0cedd16d1bce}), yet neither LDP, FDO nor DOIP specifications recommends human-readable representations to be provided
\item
  Neither DOIP nor LDP require license to be expressed (RDA-R1.1-01M, RDA-R1.1-02M, RDA-R1.1-03M), yet this is crucial for re-use and machine actionability of FAIR data and metadata to be legal
\item
  Machine-understandable types, provenance and data/metadata standards (RDA-R1.1-03M RDA-R1.3-02M, RDA-R1.3-02M, RDA-R1.3-02D) are important for machine actionability, but are currently unspecified for FDOs. \cite{126uxr5pI} explores possible machine-readable FDO types, however the type systems themselves have not yet been formalised. Linked Data on the other side have too many semantic and syntactic type systems, making it difficult to write consistent clients.
\item
  Indicators for FAIR data are weak for either approach, as too much reliance is put on metadata. For instance in Linked Data, given a URL of a CSV file, what is its persistant identifier or license information? FAIR Signposting \cite{snykkm7R} can improve findability of metadata using HTTP Link relations, which enable an FDO-like overlay for any HTTP resource. In DOIP, responses for bytestreams can include the data identifier: if that is a PID (not enforced by DOIP), its metadata is accessible.
\item
  Resolving FDOs via Handle PIDs to the corresponding DOIP server is currently undefined by FDO and DOIP specifications. \texttt{0.TYPE/DOIPServiceInfo} lookup is only possible once DOIP server is known.
\end{itemize}
